Dark Control:\\ A Unified Account of Default Mode Function\\ by Markov Decision Processes

### 2017-09-09 paper/paper.tex
integrating other comments for R1

\title{Dark Control:\\
       A Unified Account of Default Mode Function\\
       by Control Theory and Reinforcement Learning}

@@

The present paper proposes a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
DMN function is recast in mathematical terms of
control theory and reinforcement learning based on Markov decision processes.
%
We argue that our formal account of DMN function naturally accommodates as special cases
the previously proposed cognitive accounts on
(1) predictive coding,
(2) semantic associations, and
(3) a ``sentinel'' role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings in animals and humans.
\end{abstract}

@@

\subsection{Towards a formal account of default mode function: higher-order control of the organism}
The human DMN is responsible for extended parts of the baseline neural activity,
which typically decreases when engaged in psychological experiments
\citep{raichle_baseline}.
The standard mode of
neural information maintenance and manipulation has
been argued to mediate evolutionarily conserved functions
\citep{brown1914nature, binder1999conceptual, buzsaki2006rhythms}.
Today, many psychologists and neuroscientists believe that
the DMN implements some form of
probabilistic estimation of past, hypothetical, and
future events
\citep{fox2005, hassabis2007patients, schacter2007remembering, binder2009,
randy2008, spreng2009common}.
This brain network
might have emerged to continuously predict the environment using
mental imagery as an evolutionary advantage \citep{suddendorf2007evolution}.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default mode regions \citep{weiss2006}
and increased DMN activity was linked to more task-independent,
yet sometimes useful thoughts
\citep{mason2007, seli2016mind}.
%
Gaining insight into DMN function is
particularly challenging because
this network appears to
simultaneously modulate perception-action cycles and
entertain probabilistic contemplations
across time, space, and content domains
\citep{boyer2008evolutionary}.

@@

We argue that DMN implication in many
advanced cognitive processes in humans
can be recast as prediction error minimization
based on probabilistic mental simulations,
thus maximizing action outcome across multiple time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling
\citep{tenenbaum2011grow}.
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide blocks of pseudo-experience to iteratively
optimize the behavior of the organism.
%
It has indeed been proposed that
the human brain's energy budget is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
This idea is invigorated by empirical evidence from
neuroscience experiments \citep{kording2004bayesian, fiser2004small}.
The present paper proposes a
process model that satisfies this contention.
We also contribute to the discussion of DMN function
by providing some of the first empirical evidence that
morphological variability in DMN nodes
is linked to the reward circuitry (Fig. 2).
Finally, we detail how our process model relates
to previous accounts of DMN function and
we derive explicit hypotheses to be tested in future neuroscience
experiments (Box 1).

@@

\item Experiment (Animals): We hypothesize that experience replay for browsing problem solutions subserved by the DMN is necessary for choice behavior in mice. Hippocampal single-cell recordings have shown that neural patterns during experimental choice behavior are reiterated during sleep and before make analogous choices in the future. Necessity of the DMN, in addition to the hippocampus, for “mind-searching” actions during choice behavior can be demonstrated by causal dysruption of DMN nodes, such as circumscribed brain lesion or optogenetic intervention in the inferior parietal and prefrontal cortices.

@@

\section{Known neurobiological properties of the default mode network}
We begin by a neurobiological deconstruction of the DMN
based on experimental findings in the neuroscience literature.
This walkthrough across each node of the DMN will
outline the individual functional profiles,
paving the way for their algorithmic interpretation
in our formal account (section 3).

@@

Normal and disturbed metabolic fluctuations in the
human PMC have been closely related to
changes of conscious awareness \citep{cavanna2006precuneus}.
Indeed,
the PMC matures relatively late (i.e., myelination) during postnatal development in monkeys
\citep{goldman1987development}, which is generally considered to
be a sign of evolutionary sophistication.
%
This DMN node has long been speculated to
reflect constant computation of
environmental statistics and its internal representation
as an inner ``mind's eye'' \citep{cavanna2006precuneus, leech_pcc2014}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from medial damage in the parietal cortex
\citep{balint1909seelenlahmung}.
Neurological patients are plagued by an
inability to combine various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as an inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in gathering
information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity analyses
\citep{bzdok2015subspecialization}
with the amygdala, involved in significance evaluation, and
the nucleus accumbens (NAc), involved in reward evaluation.
Specifically, among all parts of the PMC,
the ventral posterior cingulate cortex was
most connected to the laterobasal
nuclei group of the amygdala
\citep{bzdok2015subspecialization}.
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological relevance assessment
\citep{amygdala_bzdok}.

The putative role of the PMC in continuous abstract integration of
environmental relevance
and ensuing top-level guidance of action on the environment is supported
by many neuroscience experiments.
Electrophysiological recordings in animals implicated PMC neurons in
strategic decision making \citep{pearson2009neurons},
risk assessment \citep{mccoy2005risk},
outcome-dependent behavioral modulation \citep{hayden2009electrophysiological},
as well as approach-avoidance behavior
\citep{vann2009does}.
Neuron spiking activity in the PMC allowed distinguishing
whether a monkey would pursue an exploratory or exploitative
behavioral strategy during food foraging \citep{pearson2009neurons}.
Further, single-cell recordings in the monkey PMC
demonstrated this brain region's sensitivity to
subjective target utility \citep{mccoy2005risk} and integration
across individual decision-making instances \citep{pearson2009neurons}.
This DMN node encoded the
preference for or aversion to options with uncertain reward outcomes
and its spiking activity was more associated with
subjectively perceived relevance of a chosen object
than by its actual value,
based on an ``internal currency of value'' \citep{mccoy2005risk}.
In fact, direct stimulation of PMC neurons
promoted exploratory actions,
which would otherwise be shunned \citep{hayden2008stim}.
Graded changes in firing rates of PMC neurons
indicated changes in upcoming choice trials, while their neural patterns were
distinct from neuronal spike firings that indicated choosing either option.
Similarly in humans,
the DMN has been shown to gather and integrate information
over different parts of auditory narratives in an fMRI study
\citep{simony2016dynamic}.

@@

The dmPFC may subserve inference, representation, and assessment
of one's own and other individuals' action considerations.
Generally,
neurological patients with tissue damage in the prefrontal cortex
are known to struggle with
adaptation to new stimuli and events
\citep{stuss1986frontal}.
Specifically, neural activity in the human dmPFC
reflected expectations about other peoples' actions and
outcomes of these predictions.
Neural activity in the dmPFC indeed explained the performance decline
of inferring other peoples' thoughts in aging humans \citep{moran2012social}.
Certain dmPFC neurons in macaque monkeys exhibited a preference
for processing others', rather than own, behavior
with fine-grained adjustment of contextual circumstances \citep{yoshida2010neural}.
% In fact, the spatially neighboring dorsal anterior cingulate cortex
% has also been linked to computing values and efforts of
% persisting a behavioral plan versus switching the
% environmental context in several lesion studies \citep{kolling2016value}.
%
Such highly abstract neural computations necessarily rely on the
construction of probabilistic internal information drawing from
memory recall, mental scene elaboration processes,
and stored knowledge of world regularities.
\begin{wrapfigure}{r}{.4\linewidth}
  \includegraphics[width=1.\linewidth]{fig_dmn.png}
  \centering
  \caption{\textbf{Morphological coupling between reward system and
  default mode network.} Based on 9,932 human subjects from the UK Biobank, inter-individual differences in left NAc volume ($R^2=0.11+/-0.02$) and right NAc volume ($R^2=0.14+/-0.02$) could be predicted from volume in the DMN nodes. These out-of-sample generalization performances were obtained from support vector regression applied to normalized node volumes in the DMN in a 10-fold cross-validation procedure.
  Consistent for the left and right reward system, NAc volume in a given subject is
  positively coupled with the vmPFC and HC.
  Code and data for reproduction and visualization:
  \url{www.github.com/banilo/to_be_added_later}.
  }
  \label{fig:dmn}
\end{wrapfigure}
%
Moreover,
in a computational neuroimaging experiment,
dorsomedial PFC activity preferentially processed the consequences of
action choices that were considered but not actually executed,
whereas ventromedial PFC (vmPFC) activity
processed especially value outcomes of actions actually performed on the environment
\citep{nicolle2012agent}.

Comparing to the dmPFC,
the vmPFC probably subserves
subjective value evaluation and
risk estimation of relevant environmental stimuli
(Fig. \ref{fig:dmn}).
The ventromedial prefrontal DMN may
subserve adaptive behavior by bottom-up-driven
processing of “what matters now”,
probably drawing on sophisticated value representations
\citep{rolls_OFC, doherty2015structure}.
Quantitative lesion findings across 344 human individuals confirmed
a substantial impairment in value-based action choice
\citep{glascher2012lesion}.
Indeed,
this DMN node is preferentially connected with reward-related and limbic regions.
The vmPFC is well known to have direct connections
with the NAc
in axonal tracing studies in monkeys \citep{haber1995orbital}.
Congruently, the gray-matter volume of the vmPFC and NAc
correlated with indices of value-guided behavior and reward attitudes
in humans
\citep{lebreton2009automatic}.
NAc activity is thought to reflect reward prediction signals
from dopaminergic neurotransmitter pathways
\citep{schultz1998predictive}
that not only channel action towards basic survival needs but also
enable more abstract reward processings, and thus perhaps RL,
in humans \citep{doherty2015structure}.
Consistently, diffusion MRI tractography in humans and monkeys
\citep{croxson2005quantitative}
quantified the NAc to
be more connected to the vmPFC than dmPFC in both species.
Two different functional connectivity analyses in humans also strongly connected
the vmPFC with the NAc, hippocampus (HC),
and PMC \citep{bzdok2015subspecialization}.
%
In line with these connectivity findings in animals and humans,
the vmPFC is often proposed to represent triggered
emotional and motivational states \citep{damasio1996somatic}.
Such real or imagined arousal states could be mapped in the vmPFC
as a bioregulatory disposition influencing cognition
and decision making.
In neuroeconomic studies of human decision making,
the vmPFC consistently reflects an individual’s subjective
value estimates
\citep{behrens2008associative}.
This may be why performance within and across participants
was related to state encoding in the vmPFC \citep{Schuck2016}.
Such a ``cognitive map'' of the action space was argued to encode
the current task state even when states are unobservable from
the sensory environment.

@@

There is hence increasing evidence that
HC function extends beyond simple forms of
encoding and reconstruction of memory and space information.
Based on spike recordings of hippocampal neuronal populations,
complex spiking patterns can be followed across extended periods including
their modification of input-free self-generated patterns
after environmental events \citep{buzsaki2004large}.
Specific spiking sequences, which were elicited by experimental task design,
have been shown to be re-enacted spontaneously during
quiet wakefulness and sleep \citep{hartley2014space, o2010play}.
Moreover, neuronal spike sequences measured in hippocampal place cells of rats
featured re-occurrence directly after experimental trials
as well as directly before upcoming experimental trials \citep{diba2007forward}.
Similar spiking patterns in hippocampal neurons during rest and sleep
have been proposed to be critical in communicating local information
to the neocortex for long-term storage, potentially also in the nodes of the DMN.
Moreover, in mice, invasively triggering spatial experience recall
in the HC during sleep
has been demonstrated to subsequently alter action choice during wakefulness
\citep{lavilleon2015}.
These HC-subserved mechanisms
probably contribute to advanced cognitive processes that require
re-experiencing or newly constructed mental scenarios,
such as in recalling autobiographical memory episodes
\citep{hassabis2007patients}.

@@

The TPJ in the right-hemispheric DMN (RTPJ)
has been shown to be closely related to
multi-sensory prediction and prediction error signaling.
It is probably central for
action initiation during goal-directed psychological tasks and for
sensorimotor behavior by integrating multi-sensory attention
\citep{corbetta2002control}.
Involvement of this DMN node was repeatedly reported in
multi-step action execution \citep{hartmann2005takes},
visuo-proprioceptive conflict \citep{Balslev2005}, and
detection of environmental changes across
visual, auditory, or tactile stimulation
\citep{downar2000multimodal}.
Direct electrical stimulation of the human
RTPJ during neurosurgery was associated with altered perception
and stimulus awareness \citep{blanke2002neuropsychology}.
%
It was argued that the RTPJ encodes actions and ensuing outcomes,
without necessarily relating those to value estimation
\citep{liljeholm2013neural, hamilton2008action,
jakobs2009effects}.
Additionally, neural activity in the RTPJ has been proposed to reflect
stimulus-driven attentional reallocation to
self-relevant and unexpected sources of information
as a ‘‘circuit breaker’’ that recalibrates control and maintenance brain networks
\citep{bzdok2013tpj, corbettashul2008}.
% (Bzdok et al., 2013; Corbetta et al., 2008).
Indeed, neurological patients with RTPJ damage have particular difficulties
with multi-step actions \citep{hartmann2005takes}.
In the face of large discrepancies between actual and previously predicted
environmental events the RTPJ acts as a potential switch between
externally-oriented mind sets focussed on the
sensory environment and internally-oriented mind sets focussed
on mental scene construction.
For instance, temporally induced RTPJ damage in humans diminished the
impact of predicted intentions of other individuals
\citep{young2010disruption},
a capacity believed to be enabled by the DMN.
The RTPJ might hence be an important relay that shifts away
from the ‘‘internally directed’’ baseline processes
to, instead, deal with unexpected environmental stimuli and events.
% (TODO: slow endogeneous changes in state transitions by the policy and value matrices)

@@

\section{Reinforcement learning: a process model for DMN function}
We now argue the outlined neurobiological properties
of the DMN nodes
to be sufficient for implementing all components
of a full-fledged RL system.
Recalling past experience, considering candidate actions,
random sampling of possible experiences, as well as
estimation of instantaneous and expected delayed reward outcomes
are key components of intelligent RL agents
that are plausible to functionally intersect in the DMN.

@@

\subsubsection{Accumulated rewards and policies}
The behavior of the agent is governed by a \textit{policy}, which maps states of the world
to probability distributions over actions.
Starting at time $t=0$,
following a policy $\pi$ generates a trajectory of action choices as follows:
\begin{eqnarray*}
  \begin{split}
    \textbf{choose action: }a_0 &\sim \pi(a|s_0)\\
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect reward }R_0 = r(s_0, a_0, s_1)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect reward }R_1 = r(s_1, a_1,s_2)\\
    \vdots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and collect reward }R_{t} = r(s_{t}, a_{t},s_{t+1})\\
    \vdots
  \end{split}
\end{eqnarray*}
We assume time-invariance
in that we expect the dynamics of the process
to be equivalent over sufficiently long time windows of equal length (i.e., stationarity).
Since an action executed in the present moment might have repercussions in the far future, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
\textit{cumulative reward} estimate which takes into account expected reward
from action choices in the future.
A common approach to
modeling this accumulation is the time-discounted cumulative reward% function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t = R_0 + \gamma R_1 + \gamma^2 R_2 + \ldots + \gamma^tR_t + \ldots
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following an action policy $\pi$.
%
Note that value buffering may be realized in the vmPFC.
This DMN node has direct connections to
to the NAc, known to be involved in reward evaluation.

@@

\pagebreak[4]
\begin{mdframed}
\hspace{2mm}
\vspace{2mm}
% neuroscience of Bellman equation
\textbf{Box 2: Neural correlates of the Bellman equation in the DMN}
Relating decomposition of consecutive action choices by the Bellman equation
to neuroscience,
specific neural activity in the dorsal prefrontal cortex (BA9)
was linked to processing ``goal-tree sequences''
in human neuroimaging experiments
\citep{koechlin1999role, koechlin2000dissociating}.
Sub-goal exploration may require multi-task switching
between cognitive processes
as later parts of a solution
frequently depend on respective earlier steps in a given solution path, which
necessitates storage of expected intermediate outcomes.
As such,
``cognitive branching'' operations for nested processing of behavioral strategies
are likely to
entail secondary reallocation of attention and working-memory resources.
Further neuroimaging experiments \citep{braver2002role} corroborated
the prefrontal DMN to subserve
``processes related to the management and monitoring of sub-goals while maintaining information in working memory''.
%
Moreover,
neurological patients with lesions in this DMN node were reported
to be impaired in aspects of realizing ``multiple sub-goal scheduling''
\citep{burgess2000cognitive}.
Hence,
the various advanced human mental abilities subserved by the DMN, such as
planning and abstract reasoning, can be viewed to involve some form of
action-decision branching to enable higher-order executive control.
\end{mdframed}

>>>>

\title{Dark Control:\\
       A Unified Account of Default Mode Function\\
       by Markov Decision Processes}

@@

The present paper proposes a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
DMN function is recast in terms of
control theory and reinforcement learning by basing it on Markov decision processes.
%
We argue that our formal account of DMN function naturally accommodates as special cases
the previously proposed cognitive accounts on
(1) predictive coding,
(2) semantic associations, and
(3) a ``sentinel'' role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings in animals and humans.
\end{abstract}

@@

\subsection{Towards a formal account of default mode function: higher-order control of the organism}
The network nodes that compose the human DMN
are responsible for extended parts of the baseline neural activity,
which typically decreases when engaged in psychological experiments
\citep{raichle_baseline}.
The standard mode of
neural information maintenance and manipulation has
been argued to mediate evolutionarily conserved functions
\citep{brown1914nature, binder1999conceptual, buzsaki2006rhythms}.
Today, many psychologists and neuroscientists believe that
the DMN implements some form of
probabilistic estimation of past, hypothetical, and
future events
\citep{fox2005, hassabis2007patients, schacter2007remembering, binder2009,
randy2008, spreng2009common}.
This brain network
might have emerged to continuously predict the environment using
mental imagery as an evolutionary advantage \citep{suddendorf2007evolution}.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default mode regions \citep{weiss2006}
and increased DMN activity was linked to more task-independent,
yet sometimes useful thoughts
\citep{mason2007, seli2016mind}.
%
Gaining insight into DMN function is
particularly challenging because
this network appears to
simultaneously modulate perception-action cycles and
entertain probabilistic contemplations
across time, space, and content domains
\citep{boyer2008evolutionary}.

@@

We argue that DMN implication in many
advanced cognitive processes in humans
can be recast as prediction error minimization
based on probabilistic mental simulations,
thus maximizing action outcome across multiple time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling
\citep{tenenbaum2011grow}.
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide blocks of pseudo-experience to iteratively
optimize the behavior of the organism.
%
It has indeed been proposed that
the human brain's energy budget is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
This idea is invigorated by empirical evidence from
neuroscience experiments \citep{kording2004bayesian, fiser2004small}.
The present paper proposes a
process model that satisfies this contention.
We also contribute to the discussion of DMN function
by providing some of the first empirical evidence that
morphological variability in DMN regions
is linked to the reward circuitry (Fig. 2).
Finally, we detail how our process model relates
to previous accounts of DMN function and
we derive explicit hypotheses to be tested in future neuroscience
experiments (Box 1).

@@

\item Experiment (Animals): We hypothesize that experience replay for browsing problem solutions subserved by the DMN is necessary for choice behavior in mice. Hippocampal single-cell recordings have shown that neural patterns during experimental choice behavior are reiterated during sleep and before make analogous choices in the future. Necessity of the DMN, in addition to the hippocampus, for “mind-searching” actions during choice behavior can be demonstrated by causal dysruption of DMN regions, such as circumscribed brain lesion or optogenetic intervention in the inferior parietal and prefrontal cortices.

@@

\section{Known neurobiological properties of the default mode network}
We begin by a neurobiological deconstruction of the DMN
based on experimental findings in the neuroscience literature.
This walkthrough across each region of the DMN will
outline the individual functional profiles,
paving the way for their algorithmic interpretation
in our formal account (section 3).

@@

Normal and disturbed metabolic fluctuations in the
human PMC have been closely related to
changes of conscious awareness \citep{cavanna2006precuneus}.
Indeed,
the PMC matures relatively late (i.e., myelination) during postnatal development in monkeys
\citep{goldman1987development}, which is generally considered to
be a sign of evolutionary sophistication.
%
This DMN region has long been speculated to
reflect constant computation of
environmental statistics and its internal representation
as an inner ``mind's eye'' \citep{cavanna2006precuneus, leech_pcc2014}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from medial damage in the parietal cortex
\citep{balint1909seelenlahmung}.
Neurological patients are plagued by an
inability to combine various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as an inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in gathering
information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity analyses
\citep{bzdok2015subspecialization}
with the amygdala, involved in significance evaluation, and
the nucleus accumbens (NAc), involved in reward evaluation.
Specifically, among all parts of the PMC,
the ventral posterior cingulate cortex was
most connected to the laterobasal
nuclei group of the amygdala
\citep{bzdok2015subspecialization}.
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological relevance assessment
\citep{amygdala_bzdok}.

The putative role of the PMC in continuous abstract integration of
environmental relevance
and ensuing top-level guidance of action on the environment is supported
by many neuroscience experiments.
Electrophysiological recordings in animals implicated PMC neurons in
strategic decision making \citep{pearson2009neurons},
risk assessment \citep{mccoy2005risk},
outcome-dependent behavioral modulation \citep{hayden2009electrophysiological},
as well as approach-avoidance behavior
\citep{vann2009does}.
Neuron spiking activity in the PMC allowed distinguishing
whether a monkey would pursue an exploratory or exploitative
behavioral strategy during food foraging \citep{pearson2009neurons}.
Further, single-cell recordings in the monkey PMC
demonstrated this brain region's sensitivity to
subjective target utility \citep{mccoy2005risk} and integration
across individual decision-making instances \citep{pearson2009neurons}.
This DMN region encoded the
preference for or aversion to options with uncertain reward outcomes
and its spiking activity was more associated with
subjectively perceived relevance of a chosen object
than by its actual value,
based on an ``internal currency of value'' \citep{mccoy2005risk}.
In fact, direct stimulation of PMC neurons
promoted exploratory actions,
which would otherwise be shunned \citep{hayden2008stim}.
Graded changes in firing rates of PMC neurons
indicated changes in upcoming choice trials, while their neural patterns were
distinct from neuronal spike firings that indicated choosing either option.
Similarly in humans,
the DMN has been shown to gather and integrate information
over different parts of auditory narratives in an fMRI study
\citep{simony2016dynamic}.

@@

The dmPFC may subserve inference, representation, and assessment
of one's own and other individuals' action considerations.
Generally,
neurological patients with tissue damage in the prefrontal cortex
are known to struggle with
adaptation to new stimuli and events
\citep{stuss1986frontal}.
Specifically, neural activity in the human dmPFC
reflected expectations about other peoples' actions and
outcomes of these predictions.
Neural activity in the dmPFC indeed explained the performance decline
of inferring other peoples' thoughts in aging humans \citep{moran2012social}.
Certain dmPFC neurons in macaque monkeys exhibited a preference
for processing others', rather than own, behavior
with fine-grained adjustment of contextual circumstances \citep{yoshida2010neural}.
% In fact, the spatially neighboring dorsal anterior cingulate cortex
% has also been linked to computing values and efforts of
% persisting a behavioral plan versus switching the
% environmental context in several lesion studies \citep{kolling2016value}.
%
Such highly abstract neural computations necessarily rely on the
construction of probabilistic internal information drawing from
memory recall, mental scene elaboration processes,
and stored knowledge of world regularities.
\begin{wrapfigure}{r}{.4\linewidth}
  \includegraphics[width=1.\linewidth]{fig_dmn.png}
  \centering
  \caption{\textbf{Morphological coupling between reward system and
  default mode network.} Based on 9,932 human subjects from the UK Biobank, inter-individual differences in left NAc volume ($R^2=0.11+/-0.02$) and right NAc volume ($R^2=0.14+/-0.02$) could be predicted from volume in the DMN regions. These out-of-sample generalization performances were obtained from support vector regression applied to normalized region volumes in the DMN in a 10-fold cross-validation procedure.
  Consistent for the left and right reward system, NAc volume in a given subject is
  positively coupled with the vmPFC and HC.
  Code and data for reproduction and visualization:
  \url{www.github.com/banilo/to_be_added_later}.
  }
  \label{fig:dmn}
\end{wrapfigure}
%
Moreover,
in a computational neuroimaging experiment,
dorsomedial PFC activity preferentially processed the consequences of
action choices that were considered but not actually executed,
whereas ventromedial PFC (vmPFC) activity
processed especially value outcomes of actions actually performed on the environment
\citep{nicolle2012agent}.

Comparing to the dmPFC,
the vmPFC probably subserves
subjective value evaluation and
risk estimation of relevant environmental stimuli
(Fig. \ref{fig:dmn}).
The ventromedial prefrontal DMN may
subserve adaptive behavior by bottom-up-driven
processing of “what matters now”,
probably drawing on sophisticated value representations
\citep{rolls_OFC, doherty2015structure}.
Quantitative lesion findings across 344 human individuals confirmed
a substantial impairment in value-based action choice
\citep{glascher2012lesion}.
Indeed,
this DMN region is preferentially connected with reward-related and limbic regions.
The vmPFC is well known to have direct connections
with the NAc
in axonal tracing studies in monkeys \citep{haber1995orbital}.
Congruently, the gray-matter volume of the vmPFC and NAc
correlated with indices of value-guided behavior and reward attitudes
in humans
\citep{lebreton2009automatic}.
NAc activity is thought to reflect reward prediction signals
from dopaminergic neurotransmitter pathways
\citep{schultz1998predictive}
that not only channel action towards basic survival needs but also
enable more abstract reward processings, and thus perhaps RL,
in humans \citep{doherty2015structure}.
Consistently, diffusion MRI tractography in humans and monkeys
\citep{croxson2005quantitative}
quantified the NAc to
be more connected to the vmPFC than dmPFC in both species.
Two different functional connectivity analyses in humans also strongly connected
the vmPFC with the NAc, hippocampus (HC),
and PMC \citep{bzdok2015subspecialization}.
%
In line with these connectivity findings in animals and humans,
the vmPFC is often proposed to represent triggered
emotional and motivational states \citep{damasio1996somatic}.
Such real or imagined arousal states could be mapped in the vmPFC
as a bioregulatory disposition influencing cognition
and decision making.
In neuroeconomic studies of human decision making,
the vmPFC consistently reflects an individual’s subjective
value estimates
\citep{behrens2008associative}.
This may be why performance within and across participants
was related to state encoding in the vmPFC \citep{Schuck2016}.
Such a ``cognitive map'' of the action space was argued to encode
the current task state even when states are unobservable from
the sensory environment.

@@

There is hence increasing evidence that
HC function extends beyond simple forms of
encoding and reconstruction of memory and space information.
Based on spike recordings of hippocampal neuronal populations,
complex spiking patterns can be followed across extended periods including
their modification of input-free self-generated patterns
after environmental events \citep{buzsaki2004large}.
Specific spiking sequences, which were elicited by experimental task design,
have been shown to be re-enacted spontaneously during
quiet wakefulness and sleep \citep{hartley2014space, o2010play}.
Moreover, neuronal spike sequences measured in hippocampal place cells of rats
featured re-occurrence directly after experimental trials
as well as directly before upcoming experimental trials \citep{diba2007forward}.
Similar spiking patterns in hippocampal neurons during rest and sleep
have been proposed to be critical in communicating local information
to the neocortex for long-term storage, potentially also in the regions of the DMN.
Moreover, in mice, invasively triggering spatial experience recall
in the HC during sleep
has been demonstrated to subsequently alter action choice during wakefulness
\citep{lavilleon2015}.
These HC-subserved mechanisms
probably contribute to advanced cognitive processes that require
re-experiencing or newly constructed mental scenarios,
such as in recalling autobiographical memory episodes
\citep{hassabis2007patients}.

@@

The TPJ in the right-hemispheric DMN (RTPJ)
has been shown to be closely related to
multi-sensory prediction and prediction error signaling.
It is probably central for
action initiation during goal-directed psychological tasks and for
sensorimotor behavior by integrating multi-sensory attention
\citep{corbetta2002control}.
Involvement of this DMN region was repeatedly reported in
multi-step action execution \citep{hartmann2005takes},
visuo-proprioceptive conflict \citep{Balslev2005}, and
detection of environmental changes across
visual, auditory, or tactile stimulation
\citep{downar2000multimodal}.
Direct electrical stimulation of the human
RTPJ during neurosurgery was associated with altered perception
and stimulus awareness \citep{blanke2002neuropsychology}.
%
It was argued that the RTPJ encodes actions and ensuing outcomes,
without necessarily relating those to value estimation
\citep{liljeholm2013neural, hamilton2008action,
jakobs2009effects}.
Additionally, neural activity in the RTPJ has been proposed to reflect
stimulus-driven attentional reallocation to
self-relevant and unexpected sources of information
as a ‘‘circuit breaker’’ that recalibrates functional control of brain networks
\citep{bzdok2013tpj, corbettashul2008}.
% (Bzdok et al., 2013; Corbetta et al., 2008).
Indeed, neurological patients with RTPJ damage have particular difficulties
with multi-step actions \citep{hartmann2005takes}.
In the face of large discrepancies between actual and previously predicted
environmental events the RTPJ acts as a potential switch between
externally-oriented mind sets focussed on the
sensory environment and internally-oriented mind sets focussed
on mental scene construction.
For instance, temporally induced RTPJ damage in humans diminished the
impact of predicted intentions of other individuals
\citep{young2010disruption},
a capacity believed to be enabled by the DMN.
The RTPJ might hence be an important relay that shifts away
from the ‘‘internally directed’’ baseline processes
to, instead, deal with unexpected environmental stimuli and events.
% (TODO: slow endogeneous changes in state transitions by the policy and value matrices)

@@

\section{Reinforcement learning control: a process model for DMN function}
We now argue the outlined neurobiological properties
of the DMN regions
to be sufficient for implementing all components
of a full-fledged RL system.
Recalling past experience, considering candidate actions,
random sampling of possible experiences, as well as
estimation of instantaneous and expected delayed reward outcomes
are key components of intelligent RL agents
that are plausible to functionally intersect in the DMN.

@@

\subsubsection{Accumulated rewards and policies}
The behavior of the agent is governed by a \textit{policy}, which maps states of the world
to probability distributions over actions.
Starting at time $t=0$,
following a policy $\pi$ generates a trajectory of action choices as follows:
\begin{eqnarray*}
  \begin{split}
    \textbf{choose action: }a_0 &\sim \pi(a|s_0)\\
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect reward }R_0 = r(s_0, a_0, s_1)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect reward }R_1 = r(s_1, a_1,s_2)\\
    \vdots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and collect reward }R_{t} = r(s_{t}, a_{t},s_{t+1})\\
    \vdots
  \end{split}
\end{eqnarray*}
We assume time-invariance
in that we expect the dynamics of the process
to be equivalent over sufficiently long time windows of equal length (i.e., stationarity).
Since an action executed in the present moment might have repercussions in the far future, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
\textit{cumulative reward} estimate which takes into account expected reward
from action choices in the future.
A common approach to
modeling this accumulation is the time-discounted cumulative reward% function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t = R_0 + \gamma R_1 + \gamma^2 R_2 + \ldots + \gamma^tR_t + \ldots
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following an action policy $\pi$.
%
Note that value buffering may be realized in the vmPFC.
This DMN region has direct connections to
to the NAc, known to be involved in reward evaluation.

@@

\pagebreak[4]
\begin{mdframed}
\hspace{2mm}
\vspace{2mm}
% neuroscience of Bellman equation
\textbf{Box 2: Neural correlates of the Bellman equation in the DMN}
Relating decomposition of consecutive action choices by the Bellman equation
to neuroscience,
specific neural activity in the dorsal prefrontal cortex (BA9)
was linked to processing ``goal-tree sequences''
in human neuroimaging experiments
\citep{koechlin1999role, koechlin2000dissociating}.
Sub-goal exploration may require multi-task switching
between cognitive processes
as later parts of a solution
frequently depend on respective earlier steps in a given solution path, which
necessitates storage of expected intermediate outcomes.
As such,
``cognitive branching'' operations for nested processing of behavioral strategies
are likely to
entail secondary reallocation of attention and working-memory resources.
Further neuroimaging experiments \citep{braver2002role} corroborated
the prefrontal DMN to subserve
``processes related to the management and monitoring of sub-goals while maintaining information in working memory''.
%
Moreover,
neurological patients with lesions in this DMN region were reported
to be impaired in aspects of realizing ``multiple sub-goal scheduling''
\citep{burgess2000cognitive}.
Hence,
the various advanced human mental abilities subserved by the DMN, such as
planning and abstract reasoning, can be viewed to involve some form of
action-decision branching to enable higher-order executive control.
\end{mdframed}
